\documentclass[11pt]{article}
\usepackage[a4paper, left=1.5cm,right=1.5cm,top=2cm,bottom=2cm]{geometry}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsmath}  % Math environments and symbols
\usepackage{amssymb}  % \mathbb and extra symbols
\usepackage{booktabs} % Nicer tables (\toprule, \midrule, \bottomrule)

\title{Compositional Latent Space}
\author{Suresh Murugaiyan}
\date{July 2026}

\begin{document}

\maketitle

\section{Introduction}

We train a \emph{compositional autoencoder} (Path~A of the compositional
operator-network framework) on the FlowBench 2D lid-driven cavity dataset:
steady Navier--Stokes flow with an object placed inside the cavity. The
autoencoder compresses each steady flow field into a small latent vector
$\boldsymbol{z}$ and reconstructs the field from it --- a nonlinear analogue
of POD. The key structural choice is that $\boldsymbol{z}$ is not one opaque
vector; it is partitioned by architecture into three named blocks, each
intended to carry one physical factor:
%
\begin{equation}
  \boldsymbol{z} \;=\;
  [\, \underbrace{\boldsymbol{z}_{\mu}}_{\text{regime }(4)}
  \,\|\,
  \underbrace{\boldsymbol{z}_{g}}_{\text{geometry }(32)}
  \,\|\,
  \underbrace{\boldsymbol{z}_{\xi}}_{\text{residual }(16)} \,]
  \;\in\; \mathbb{R}^{52}.
\end{equation}
%
The hypothesis under test: after training, $\boldsymbol{z}_{\mu}$ should
contain the Reynolds number, $\boldsymbol{z}_{g}$ should contain the object
shape, and neither should leak into the other. Because the flow is steady,
there is no dynamics block $\boldsymbol{z}_{\eta}$ and no latent propagator
$\Phi$ in this first version; those enter later with time-dependent cases.

\section{Dataset}

Each FlowBench sample is a steady lid-driven cavity solution at
$512 \times 512$ resolution. The data loader
(\texttt{data/dataset.py}) unpacks the raw \texttt{.npz} arrays into:
%
\begin{itemize}
  \item \textbf{fields} --- the $(u, v, p)$ solution, resized
        $512^2 \to 256^2$ to keep training cheap;
  \item \textbf{sdf} --- the signed distance field of the object
        (the geometry label);
  \item \textbf{mask} --- binary fluid mask ($1$ in fluid, $0$ in the solid);
  \item \textbf{log\_re} --- the regime label. Re is stored as a constant
        channel, so a single pixel is extracted, transformed to
        $\log_{10}\mathrm{Re}$ (Re spans orders of magnitude), and
        standardized with \emph{train-set} statistics; the test set reuses
        those statistics so both splits share one scale.
\end{itemize}
%
The loader also computes three cheap per-sample geometry summaries from the
mask (solid area fraction, centroid $x$, centroid $y$). These are never used
during training; they serve only as ground-truth targets for the linear-probe
diagnostic of Section~\ref{sec:diagnostics}.

\section{Network architecture}

Four modules (\texttt{models/compositional/networks.py}):
%
\begin{itemize}
  \item \textbf{FieldEncoder} $\mathcal{E}$ --- a CNN that halves the spatial
        resolution repeatedly ($256 \to 8$), pools, and terminates in
        \emph{three separate linear heads} producing
        $\boldsymbol{z}_{\mu}, \boldsymbol{z}_{g}, \boldsymbol{z}_{\xi}$.
        The blocks are architecturally separate from the start.
  \item \textbf{FieldDecoder} $\mathcal{D}$ --- the mirror image: the
        concatenated $\boldsymbol{z}$ is reshaped to an $8\times 8$ feature
        map and upsampled back to a $256^2$ $(u, v, p)$ field.
  \item \textbf{RegimeHead} --- a small MLP that reads
        $\boldsymbol{z}_{\mu}$ \emph{only} and must predict
        $\log \mathrm{Re}$.
  \item \textbf{SDFHead} --- a small decoder that reads $\boldsymbol{z}_{g}$
        \emph{only} and must redraw the object's SDF at $64 \times 64$.
\end{itemize}
%
The two heads are the supervision mechanism: since the regime head sees only
$\boldsymbol{z}_{\mu}$, the encoder is forced to route Reynolds-number
information into that block, and likewise shape information into
$\boldsymbol{z}_{g}$.

\section{Training objective}

Each optimization step combines four losses
(\texttt{models/compositional/compositional\_ae.py}), weighted by
$\lambda$-coefficients set in the configuration file:
%
\begin{enumerate}
  \item \textbf{Masked reconstruction (L1).} MSE between decoded and true
        fields, evaluated over fluid points only: predictions inside the
        solid are zeroed and the error is normalized by the fluid node
        count, so samples with large objects are not unfairly weighted.
  \item \textbf{Regime supervision.} MSE of
        $\mathrm{RegimeHead}(\boldsymbol{z}_{\mu})$ against the standardized
        $\log_{10}\mathrm{Re}$. Pulls Re into $\boldsymbol{z}_{\mu}$.
  \item \textbf{Geometry supervision.} MSE of
        $\mathrm{SDFHead}(\boldsymbol{z}_{g})$ against the downsampled true
        SDF. Pulls shape into $\boldsymbol{z}_{g}$.
  \item \textbf{Cross-block decorrelation (L6).} The mean absolute Pearson
        correlation, computed across the batch, between every coordinate of
        one block and every coordinate of another, penalized for all three
        block pairs. Losses~2--3 pull information \emph{into} the correct
        block; this term pushes redundant copies \emph{out of} the wrong
        blocks.
\end{enumerate}
%
The residual block $\boldsymbol{z}_{\xi}$ receives no supervision at all: it
is free capacity that absorbs whatever the reconstruction needs beyond
$(\mathrm{Re}, \text{shape})$, which is exactly the role of the residual
block in the framework.

Training is driven by \texttt{main.py}: it loads the YAML configuration,
seeds all libraries, builds the train/test datasets (passing the train
Re-statistics to the test set), and trains with PyTorch Lightning,
checkpointing on the best validation reconstruction error.

\section{Linear-probe diagnostics}
\label{sec:diagnostics}

The diagnostic (\texttt{diagnostics/probes.py}) is the actual scientific
test. After training, the entire test set is encoded, and for each
(block, target) pair a ridge regression is fitted with 5-fold
cross-validation; the resulting $R^2$ values are tabulated. For a
compositional latent we expect the block-diagonal pattern of
Table~\ref{tab:probes}.

\begin{table}[htbp]
  \centering
  \caption{Expected linear-probe $R^2$ pattern for a compositional latent
  space. High $R^2$ on the matching block and low $R^2$ elsewhere indicates
  a successful factorization; a high off-diagonal entry indicates leakage.}
  \label{tab:probes}
  \begin{tabular}{lccc}
    \toprule
    Target & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$    & \textbf{high} & low           & low \\
    Area fraction         & low           & \textbf{high} & low \\
    Centroid $x$          & low           & \textbf{high} & low \\
    Centroid $y$          & low           & \textbf{high} & low \\
    \bottomrule
  \end{tabular}
\end{table}

If the trained latent reproduces this pattern, the factorization worked:
Re is linearly readable from the regime block and \emph{not} from the
geometry block, and vice versa. A high $R^2$ in an off-diagonal cell means
information leaked --- exactly what the planned loss ablations are designed
to study (e.g., does removing the decorrelation penalty cause leakage?).

\section{Results}

\subsection{Run 1: baseline loss weights}

First training run on the FlowBench LDC ``easy'' split (2{,}400 train /
600 test samples): 200 epochs, batch size 16, resolution $256^2$, loss
weights $\lambda_{\mathrm{recon}} = 1.0$, $\lambda_{\mathrm{regime}} =
\lambda_{\mathrm{geo}} = 0.1$, $\lambda_{\mathrm{decorr}} = 0.01$
(single NVIDIA GPU, $\sim$18 minutes). Final masked reconstruction error:
train $2.7 \times 10^{-3}$, validation $2.1 \times 10^{-3}$, with no sign
of overfitting. Table~\ref{tab:probes-run1} reports the measured
linear-probe $R^2$.

\begin{table}[htbp]
  \centering
  \caption{Run 1: measured linear-probe $R^2$ on the 600-sample test split
  ($\lambda_{\mathrm{decorr}} = 0.01$). Bold marks the block that should
  carry each target.}
  \label{tab:probes-run1}
  \begin{tabular}{lccc}
    \toprule
    Target & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$    & \textbf{0.994} & 0.900          & 0.777 \\
    Area fraction         & 0.042          & \textbf{0.688} & 0.172 \\
    Centroid $x$          & 0.027          & \textbf{0.736} & 0.373 \\
    Centroid $y$          & $-0.013$       & \textbf{0.690} & 0.214 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{What worked.} The regime block is clean:
$\boldsymbol{z}_{\mu}$ reads $\log \mathrm{Re}$ at $R^2 = 0.994$ and
contains essentially no geometry ($R^2 \le 0.04$). Geometry is readable
from $\boldsymbol{z}_{g}$ ($R^2 \approx 0.69$--$0.74$) and nearly absent
from $\boldsymbol{z}_{\mu}$. The moderate (rather than high) geometry
recovery is not alarming: the probes are linear, while area and centroid
are nonlinear functionals of the SDF that $\boldsymbol{z}_{g}$ encodes.

\paragraph{What failed.} Reynolds-number information leaked into the
other blocks: $\log \mathrm{Re}$ is readable from $\boldsymbol{z}_{g}$ at
$R^2 = 0.900$ and from $\boldsymbol{z}_{\xi}$ at $0.777$. This is exactly
the factor-scrambling that the impossibility result of Locatello et al.\
predicts for unconstrained latents: the decoder needs Re to reconstruct
the field and is free to stash copies of it in every block, and the
reconstruction loss rewards doing so. The decorrelation penalty is meant
to prevent this, but at $\lambda_{\mathrm{decorr}} = 0.01$ it is roughly
two orders of magnitude weaker than the reconstruction term and loses the
competition.

\paragraph{Next experiment.} Since the leaked Re is \emph{linearly}
recoverable from $\boldsymbol{z}_{g}$ --- which is precisely what a
Pearson penalty can act on --- the next run raises
$\lambda_{\mathrm{decorr}}$ from $0.01$ to $0.5$ with all else fixed.
Success criteria: the off-diagonal $\log \mathrm{Re}$ entries drop
substantially (toward $R^2 \lesssim 0.3$) while the reconstruction error
and the $\boldsymbol{z}_{g}$ geometry row remain near their Run-1 values.
If a strong Pearson weight cannot remove the leakage, the natural
escalation is HSIC (L7, nonlinear dependence) and the swap-consistency
loss (L12), which makes the factorization functional rather than merely
statistical.

\subsection{Run 2: strong Pearson decorrelation}

Identical to Run~1 except $\lambda_{\mathrm{decorr}} = 0.5$ (a $50\times$
increase). Table~\ref{tab:probes-run2} reports the measured linear-probe
$R^2$, with the Run-1 values for comparison.

\begin{table}[htbp]
  \centering
  \caption{Run 2: measured linear-probe $R^2$ on the test split
  ($\lambda_{\mathrm{decorr}} = 0.5$), with Run-1 values
  ($\lambda_{\mathrm{decorr}} = 0.01$) in parentheses. Bold marks the
  block that should carry each target.}
  \label{tab:probes-run2}
  \begin{tabular}{lccc}
    \toprule
    Target & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$ & \textbf{0.886} (0.994) & 0.719 (0.900) & 0.487 (0.777) \\
    Area fraction      & $-0.002$ (0.042)       & \textbf{0.402} (0.688) & 0.262 (0.172) \\
    Centroid $x$       & $-0.016$ (0.027)       & \textbf{0.565} (0.736) & 0.378 (0.373) \\
    Centroid $y$       & 0.016 ($-0.013$)       & \textbf{0.427} (0.690) & 0.115 (0.214) \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Outcome: a Pareto trade-off, not a fix.} The leakage shrank
--- $\log \mathrm{Re}$ from $\boldsymbol{z}_{g}$ dropped from $0.900$ to
$0.719$ and from $\boldsymbol{z}_{\xi}$ from $0.777$ to $0.487$ --- but
so did everything else: the on-diagonal regime readout degraded
($0.994 \to 0.886$) and the geometry recovery from $\boldsymbol{z}_{g}$
fell by roughly $0.2$ across all three descriptors. The strong Pearson
penalty cannot distinguish redundant copies of information (which we want
removed) from informative, structured coordinates (which we want kept),
so at large weight it partially scrambles all blocks while still leaving
the leakage far above the $R^2 \lesssim 0.3$ target. A $50\times$
increase in decorrelation pressure bought a modest leakage reduction at a
real cost --- strong evidence that weight tuning of a statistical
independence penalty alone will not functionally separate the blocks.

\paragraph{Next experiments.} Two directions, in order:
(i)~$\lambda_{\mathrm{decorr}} = 0.1$ as an intermediate sweep point ---
together with Runs~1--2 this begins the (reconstruction,
identifiability) Pareto front that the project plan targets;
(ii)~\emph{same-factor invariance} (L10): the dataset contains each
geometry at multiple Reynolds numbers, so we can penalize the variance of
$\boldsymbol{z}_{g}$ across samples that share a geometry. This demands
directly that $\boldsymbol{z}_{g}$ be unchanged when only Re changes ---
a functional constraint on the factorization rather than a statistical
one, and the first step toward the group-structured supervision at the
heart of the framework's identifiability argument.

\subsection{Run 3: intermediate decorrelation --- the Pareto verdict}

Identical to Runs~1--2 except $\lambda_{\mathrm{decorr}} = 0.1$.
Table~\ref{tab:probes-run3} completes the three-point weight sweep.

\begin{table}[htbp]
  \centering
  \caption{Pearson-decorrelation weight sweep: linear-probe $R^2$ on the
  test split for $\lambda_{\mathrm{decorr}} \in \{0.01, 0.1, 0.5\}$
  (Runs 1, 3, 2). Bold marks the block that should carry each target.}
  \label{tab:probes-run3}
  \begin{tabular}{llccc}
    \toprule
    Target & $\lambda_{\mathrm{decorr}}$ & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$ & 0.01 & \textbf{0.994} & 0.900 & 0.777 \\
                       & 0.1  & \textbf{0.978} & 0.865 & 0.752 \\
                       & 0.5  & \textbf{0.886} & 0.719 & 0.487 \\
    \midrule
    Area fraction      & 0.01 & 0.042    & \textbf{0.688} & 0.172 \\
                       & 0.1  & 0.017    & \textbf{0.657} & 0.368 \\
                       & 0.5  & $-0.002$ & \textbf{0.402} & 0.262 \\
    \midrule
    Centroid $x$       & 0.01 & 0.027    & \textbf{0.736} & 0.373 \\
                       & 0.1  & 0.014    & \textbf{0.712} & 0.493 \\
                       & 0.5  & $-0.016$ & \textbf{0.565} & 0.378 \\
    \midrule
    Centroid $y$       & 0.01 & $-0.013$ & \textbf{0.690} & 0.214 \\
                       & 0.1  & 0.012    & \textbf{0.657} & 0.245 \\
                       & 0.5  & 0.016    & \textbf{0.427} & 0.115 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Verdict on statistical decorrelation.} The intermediate weight
recovers nearly all of Run~1's representation quality (regime readout
$0.978$, geometry $0.66$--$0.71$) but buys almost no leakage reduction
($\log \mathrm{Re}$ from $\boldsymbol{z}_{g}$: $0.900 \to 0.865$). Across
the sweep, leakage falls only when representation quality is heavily
taxed: there is no weight at which the Pearson penalty removes the
redundant Reynolds copy while preserving the blocks' intended content.
This is a clean negative result, and it matches the framework's
prediction: statistical independence penalties cannot make a
factorization \emph{functional}. The next run therefore moves to the
structural tool.

\paragraph{Run 4: same-factor invariance (L10).} The dataset contains
each geometry at multiple Reynolds numbers, which enables group-structured
minibatches: each training batch is composed of a few geometry groups
with several Re per group, and the loss
%
\begin{equation}
  \mathcal{L}_{\mathrm{inv}}
  \;=\;
  \mathbb{E}_{\text{groups}}
  \Bigl[ \bigl\| \boldsymbol{z}_{g}^{(i)} -
  \overline{\boldsymbol{z}_{g}}^{\,\text{group}} \bigr\|^2 \Bigr]
\end{equation}
%
penalizes any variation of $\boldsymbol{z}_{g}$ within a group. Since
samples in a group differ \emph{only} in Re, any Re content in
$\boldsymbol{z}_{g}$ contributes directly to this loss: it demands
invariance as a functional property rather than decorrelation as a batch
statistic. Run 4 uses $\lambda_{\mathrm{decorr}} = 0.01$ (the Run-1
value) plus the invariance term, so that any improvement is attributable
to L10 alone.

\subsection{Run 4: same-factor invariance --- the structural loss succeeds}

Run 4 uses the Run-1 weights ($\lambda_{\mathrm{decorr}} = 0.01$) plus the
same-factor invariance term with $\lambda_{\mathrm{inv}} = 0.1$, trained
on group-structured minibatches (four geometry groups of four Reynolds
numbers per batch of 16). Table~\ref{tab:probes-run4} reports the probe
$R^2$ against the Run-1 baseline.

\begin{table}[htbp]
  \centering
  \caption{Run 4: linear-probe $R^2$ with the same-factor invariance loss
  (L10), Run-1 values in parentheses. Bold marks the block that should
  carry each target.}
  \label{tab:probes-run4}
  \begin{tabular}{lccc}
    \toprule
    Target & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$ & \textbf{0.986} (0.994) & 0.215 (0.900) & 0.412 (0.777) \\
    Area fraction      & $-0.003$ (0.042)       & \textbf{0.677} (0.688) & 0.130 (0.172) \\
    Centroid $x$       & 0.009 (0.027)          & \textbf{0.756} (0.736) & 0.143 (0.373) \\
    Centroid $y$       & 0.011 ($-0.013$)       & \textbf{0.711} (0.690) & 0.047 (0.214) \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Outcome.} The Reynolds leakage into the geometry block
collapsed from $R^2 = 0.900$ to $0.215$ --- below the $0.3$ target ---
at essentially no cost: the regime readout is intact ($0.986$) and the
geometry recovery from $\boldsymbol{z}_{g}$ is marginally \emph{better}
than the baseline ($0.68$--$0.76$). Even the residual block cleaned up
substantially ($0.777 \to 0.412$) although nothing constrains it
directly: with $\boldsymbol{z}_{g}$ forced to be Re-free, the decoder
relies on $\boldsymbol{z}_{\mu}$ for regime information and the redundant
copies decay. Contrasted with the weight sweep of
Table~\ref{tab:probes-run3} --- where a $50\times$ stronger statistical
penalty could not achieve this at any acceptable price --- this is the
framework's central claim demonstrated on real data: \emph{functional}
constraints (invariance within factor groups, enabled by the factorized
experimental design of the dataset) separate the blocks where
\emph{statistical} independence penalties cannot. The probe matrix now
shows the intended block-diagonal pattern for both supervised blocks.

\paragraph{Remaining gap and next experiments.} The residual block still
carries moderate Re content ($R^2 = 0.412$). Natural next steps:
(i)~\emph{swap consistency} (L12) --- decode $\boldsymbol{z}_{\mu}$ from
one sample with $\boldsymbol{z}_{g}$ from another and compare to the true
cross-combination, upgrading the diagnostic from ``blocks are separate''
to ``blocks are \emph{recombinable}''; (ii)~concept-vector arithmetic on
$\boldsymbol{z}_{\mu}$ and $\boldsymbol{z}_{g}$; (iii)~shrinking
$L_{\xi}$, since a steady, deterministic dataset arguably needs little
residual capacity.

\subsection{Run 5: adding swap consistency}

Run 5 adds the swap-consistency loss (L12) on top of the Run-4
configuration ($\lambda_{\mathrm{swap}} = 0.1$). For in-batch pairs
$(i, k)$ at the same Reynolds number but different geometry --- 87\% of
training samples have such a partner --- the decoder must reproduce
sample $k$'s true field from the hybrid latent
$[\boldsymbol{z}_{\mu}^{(i)} \,\|\, \boldsymbol{z}_{g}^{(k)} \,\|\,
\boldsymbol{z}_{\xi}^{(k)}]$: the regime code must be \emph{functionally
interchangeable} across geometries at fixed dynamic similarity.
Table~\ref{tab:probes-run5} reports the probes, with Run-4 values in
parentheses.

\begin{table}[htbp]
  \centering
  \caption{Run 5: linear-probe $R^2$ with invariance (L10) plus swap
  consistency (L12); Run-4 values (L10 only) in parentheses. Bold marks
  the block that should carry each target.}
  \label{tab:probes-run5}
  \begin{tabular}{lccc}
    \toprule
    Target & $\boldsymbol{z}_{\mu}$ & $\boldsymbol{z}_{g}$ & $\boldsymbol{z}_{\xi}$ \\
    \midrule
    $\log \mathrm{Re}$ & \textbf{0.972} (0.986) & 0.343 (0.215) & 0.356 (0.412) \\
    Area fraction      & $-0.005$ ($-0.003$)    & \textbf{0.693} (0.677) & 0.170 (0.130) \\
    Centroid $x$       & $-0.003$ (0.009)       & \textbf{0.767} (0.756) & 0.235 (0.143) \\
    Centroid $y$       & 0.003 (0.011)          & \textbf{0.718} (0.711) & 0.069 (0.047) \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Outcome.} The block-diagonal pattern is preserved, the
predicted cleanup of the residual block occurred (Re from
$\boldsymbol{z}_{\xi}$: $0.412 \to 0.356$; swap decoding uses the
target's $\boldsymbol{z}_{\xi}$ with the source's
$\boldsymbol{z}_{\mu}$, punishing Re content hidden in the residual),
and geometry recovery from $\boldsymbol{z}_{g}$ is the best of all runs
($0.69$--$0.77$). The one cell that moved the wrong way is Re from
$\boldsymbol{z}_{g}$ ($0.215 \to 0.343$); differences of this size
between single-seed runs should not be over-interpreted --- resolving
them requires the multi-seed protocol of the project plan. The honest
summary at this stage: \emph{same-factor invariance (L10) is the
decisive ingredient for block separation; the marginal effect of L12 on
the probe matrix is small and within plausible seed variance.}

\paragraph{What the probes cannot see.} Linear probes measure whether a
block \emph{contains} the right information, not whether the blocks are
\emph{recombinable through the decoder} --- which is precisely the
property L12 trains. The next diagnostic increment is therefore a
swap-error metric on the test set: encode two samples, decode the hybrid
latent, and compare against the true cross-combination field; reported
alongside the ordinary reconstruction error, it quantifies
compositionality directly. Further planned steps: multi-seed variance
estimates for Runs 4--5, and concept-vector arithmetic on
$\boldsymbol{z}_{\mu}$ and $\boldsymbol{z}_{g}$.

\subsection{Swap-error diagnostic: separation is not recombinability}

The swap-error metric decodes the hybrid latent
$[\boldsymbol{z}_{\mu}^{(i)} \,\|\, \boldsymbol{z}_{g}^{(k)} \,\|\,
\boldsymbol{z}_{\xi}^{(k)}]$ for same-Re, different-geometry test pairs
and compares against sample $k$'s true field; reported as a ratio to the
ordinary reconstruction error on the same samples, so ratio $\approx 1$
means swapping in another sample's regime code costs nothing. Because
the metric needs no retraining, it was evaluated retroactively on the
three key checkpoints (Table~\ref{tab:swap}).

\begin{table}[htbp]
  \centering
  \caption{Same-Re swap-consistency error on the test split (256 pairs),
  evaluated retroactively on stored checkpoints.}
  \label{tab:swap}
  \begin{tabular}{lccc}
    \toprule
    Run & Losses & Probe separation & Swap ratio \\
    \midrule
    Run 1 & reconstruction $+$ weak L6 & poor (Re everywhere) & 1.06 \\
    Run 4 & $+$ invariance (L10)       & good                 & 1.35 \\
    Run 5 & $+$ swap (L12)             & good                 & \textbf{1.04} \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Interpretation.} The three numbers tell a story the probe
tables could not. Run 1 passes the swap test (1.06) \emph{for the wrong
reason}: the decoder held redundant Re copies in every block, so it
barely depended on which sample's $\boldsymbol{z}_{\mu}$ it received.
Run 4's training removed the redundancy --- and thereby made the decoder
genuinely dependent on $\boldsymbol{z}_{\mu}$, so sample-to-sample noise
in the regime code now propagates to the output (1.35): \emph{separation
alone makes the factorization brittle}. Run 5's swap loss trains
exactly that sensitivity away, restoring ratio 1.04 while keeping the
probe separation. The combination --- blocks that are both separated
\emph{and} functionally interchangeable --- requires both structural
losses; neither the probe matrix nor the swap ratio alone certifies it.
Run 5 is the best configuration overall.

\paragraph{Cross-Re swap.} The same-Re swap is a consistency test, but a
weak one when regime supervision makes $\boldsymbol{z}_{\mu}$ nearly
deterministic in Re. The diagnostic suite therefore now also includes a
\emph{cross-Re} swap: take $\boldsymbol{z}_{\mu}$ from a sample at
$\mathrm{Re}_i$ and $\boldsymbol{z}_{g}, \boldsymbol{z}_{\xi}$ from a
different geometry observed at a different Re, decode, and compare
against the true field of that geometry at $\mathrm{Re}_i$ (present in
the dataset). This is genuine \emph{transfer} --- using the regime code
to move a geometry to a new operating point --- and cannot be passed by
Re-determinism of $\boldsymbol{z}_{\mu}$. Cross-Re ratios for the three
checkpoints are the next measurement.

\section{Next steps}

Deliberately excluded from this first version and planned as increments once
the baseline trains and the probe table looks sane:
%
\begin{itemize}
  \item swap-consistency loss (L12): decode $\boldsymbol{z}_{\mu}$ from one
        sample with $\boldsymbol{z}_{g}$ from another and compare against
        the true cross-combination;
  \item group-structured minibatches for iVAE-style identifiability;
  \item concept-vector arithmetic diagnostic;
  \item INR-style implicit decoder; HSIC (nonlinear) decorrelation;
  \item dynamics block $\boldsymbol{z}_{\eta}$ and latent propagator $\Phi$
        for time-dependent cases.
\end{itemize}

\newpage

\appendix

\section{Appendix: Running the code on Nova cluster}


\begin{itemize}
    \item STEP1: Login into the Nova cluster ssh username@nova.its.iastate.edu
    \item STEP2: Activate the virtual environment that contains all the required packages.
    \item STEP3: Request resource allocation from Nova using salloc -N 1 -n 4 -t 01:00:00 --gres=gpu:a100 --mem 369G 
    \item python main.py --config configs/compositional/conf.yaml
    \item $python diagnostics/probes.py \
  --config configs/compositional/conf.yaml \
  --checkpoint checkpoints/compositional/version_2/cae-epoch=189.ckpt$
\end{itemize}

\end{document}
